\documentclass[a4paper,11pt]{book}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Parametres du document
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\author}{Emmanuel Branlard}
\newcommand{\authorb}{E. Branlard}
\newcommand{\titlea}{Wind turbine tip-loss corrections}
\newcommand{\titleasub}{Review, implementation and investigation of new models}
\newcommand{\titleb}{\textsc{Master's Thesis}}
\renewcommand{\date}{September 2011}
\newcommand{\myyear}{2011}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Preambule
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{pre/preambule.tex}
%\input{pre/preambulePS.tex}
\input{pre/preambuleCode.tex}
\input{pre/preambuleFigs.tex}
\input{pre/preambuleMaths.tex}
\input{pre/preambuleNotes.tex}
\input{pre/preambuleNumEnv.tex}
\input{pre/preambuleTables.tex}
\input{pre/preambuleLists.tex}
\input{pre/preambuleRefIndex.tex}
\input{pre/preambuleAbr.tex}
\input{pre/preambuleLengths.tex}
\input{pre/preambuleFrontPage.tex}
\newdateformat{mydate}{\twodigit{\THEDAY}/\twodigit{\THEMONTH}/\THEYEAR}\mydate
\newcommand{\compileddate}{\today}
 %\dominitoc
 %\doparttoc
\makeindex
\makeglossary



%\newcommand{\conf}[1]{Confidential content removed. } 
\newcommand{\conf}[1]{#1 } 

\newcommand{\conffig}[1]{#1}
\newcommand{\conftext}[2]{#1}
%\newcommand{\conffig}[1]{ConfidentialFigure}
\newcommand{\confsize}[1]{#1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%Debut Document
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\pagedegarde
\mytableofcontents


\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}
\addtocontents{toc}{\protect\vspace{0.5cm}}
\myrhead{INTRODUCTION}

\paragraph{Context and motivations}
Early investigations of wind turbine performances revealed that the power produced was lower than the one expected by the blade element momentum theory, the production was as if the size of the rotor had been reduced in a proportion of approximately $3\%$. The main reason for this power loss is the circulation of flow around the tip of the blade which is a well known phenomenon in the study of aircraft wings where a vortex is emitted at the extremity of each wing. These vortices are commonly known as they can be observed under some particular pressure and humidity characteristics of the air. At the beginning of the 20$^{\text{th}}$ century, Prandtl was among the first to investigate the losses accounted by these vortices for a wing and later for a propeller. The losses appear for the former when the wing has a finite span, and for the later when the rotating devises has a finite number of blades. Prandtl's research led to the introduction of a tip-loss factor which is now widely used in wind turbine aerodynamic codes to account for the tip-losses. The load reduction at the tip due to this effect has a dramatic influence on the produced power because the tip being at a great distance to the hub it has the possibility to generate a lot of torque by a simple lever arm rule. Summing up over the entire range of operation of a wind turbine these losses can represent a reduction of 10\% on the Annual Energy Production. A proper blade design can reduce the power losses and thus increase the productivity of the blade but this require of course a better understanding of tip-losses. It is the purpose of this document to focus on this aspect.

\paragraph{Approach description} The notion of tip-losses requires for its understanding a wide picture of the different theories applied to wind turbines but also general notions of wing and wind turbines 3D aerodynamics. Historically, tip-losses appeared in the context of investigation of optimal propellers with an important contribution from authors of the German school: Prandtl, Betz, and Goldstein. Glauert in a later work applied these notions and integrated them into a wind turbine performance prediction algorithm called the Blade Element Momentum(BEM) theory. The formulations and notions of these theories will be presented in this study. A pragmatic approach will have to be taken when studying the different tip-loss functions and their implementations for they have and are still rising debates in the scientific community. These foundations established, tools like vortex codes and Computational Fluid Dynamics (CFD) will be used to investigate further the notion of tip-losses and possibly improve the performances of BEM codes.


\paragraph{Content} This document is structured with a succession of six chapters which follow the approach mentioned in the above paragraph. All chapters are progressively built upon each other, except for chapters 4 and 5 that can be read in different order. Different chapters were placed in appendices for a clearer and smoother reading, despite the fact that a large part of the corpus rely on them. In a first chapter, the different theories and three dimensional effects that are connected to tip-losses are progressively introduced. In a second time, the different tip-loss corrections found in the literature are reviewed with a focus on the main theories and their implementations. The next chapters both investigate tip-losses using a specific code. The focus of this study was to use a vortex code to derive tip-loss corrections that could be implemented in BEM codes. The study of CFD data give rise to more discussion and interesting results comparisons which will be summarized in a final chapter.


\paragraph{Contribution from this work} The different aspects on which this study contributes to existing work on the matter of tip-losses can be listed as follow:
\begin{itemize}
    \item This document is the first one to the author's knowledge to be entirely dedicated to wind turbine tip-losses. Along this line, a listing of all the different tip-loss corrections implemented was performed and important distinctions were made.  
    \item In this study, a new method to determine tip-losses using a vortex code has been described and implemented. From this method, a new tip-loss model was implemented in a BEM code in order to reproduce the 3D effects inherently present in a vortex code.
    \item A method to fit curves using the formulation of B\'ezier curves was described and developed. Such method can be widely used to describe curves such as lift, circulation or chord distribution.
    \item The author introduced the naming of ``performance tip-loss'' factor, which is a correction to the airfoil coefficients due to the tri-dimensionality of the flow at the tip. Such corrections have been investigated for instance in \cite{shen:2005} in a slightly different way. A model for the performance tip-loss function is introduced in \autoref{sec:Fclgeneral}.
    \item In \autoref{anx:goldstein}, a detailed description of Goldstein's article from 1929\cite{goldstein:1929} can be found. This article is generally considered as a complex article so that the details presented in this report is intended to help the curious reader to go quickly through this essential reference on the topic of tip-losses.
    \item In \autoref{anx:goldsteinhelical}, the method of Okulov\cite{okulov:2007} to compute Goldstein's factor at a reasonable computational cost is provided with more details than in the original article in order to share it.
    \item Eighth order polynomial that fits Goldstein's factor for different operating conditions are also provided in \autoref{sec:goldsteinfit}.
    \item The computation of Goldstein's factor being accessible, a method to use this factor in the BEM method was presented in \autoref{sec:goldsteinbem}.
    \item Analysis from \cite{hansen:2004} and \cite{johansen:2004} use CFD to investigate the axial induction. In this study similar analysis were performed with focus this time on the tip-loss factor.
    \item Along the same line than the point above, the question of definition of the local induction factor $a_B$ on the blade was risen. Different method to define it were investigated in \autoref{sec:cfdtiplossaxialinduction}
    \item All the figures are done with the wind turbine convention, rotating clockwise and with a relative wake velocity in the opposite of the stream direction. When studying the theory of helical wakes, most references uses different or inconsistent conventions.
\end{itemize}


\paragraph{About this document} This document was entirely written by the author. All the figures present in this document are from the author, either drawn or generated using Matlab or Mathematica. If a figure was inspired by others a reference is mentioned. The author borrowed figures from his own previous work, mainly\cite{branlard:2011}. 
All references cited in this document have been consulted by the author except for the following which were not available\cite{rasmussen:1988}\cite{madsen:1988}\cite{ronsten:1991}\cite{bruining:1993}\cite{rankine:1865}\cite{himmelskamp:1945}\cite{savino:1985}\cite{corrigan:1994}. This document was written in \LaTeX{} using a template from the author. The paper size chosen is A4.

\nocite{branlard:gusts}
\nocite{komninos:2008}

% \paragraph{Time line for the future months}
% \begin{itemize}
%     \item May: Compare the different methods with CFD (collaboration with the team) - Investigate the 3D corrections applied to airfoil data - Learn about Siemens BEM codes -  Suggest some AeroDyn wrapping if it seems needed - Discuss with Paul the possibility to use measurement data - Investigate options for using/implementing/improving a vortex code with Kristian and Mac who have such Matlab codes. 
%     \item June: Start seriously to use/implement a vortex code. Go-on with comparison with CFD. Maybe think about comparing with the NREL PHASE VI experiment.
%     \item July: vortex codes implementation should be done - testing should be performed before the end of the month (Elliptic wing, Goldstein's theory, Xfoil, used for validation) - Start tip-losses investigation.
%     \item August: Investigation of tip losses using the vortex code. See how this can be used for a BEM code and for blade design.  
%     \item September: Wrap-up and smooth retirement.
% \end{itemize}
% 



% \paragraph{About this work}
% variable names
% figures matfig



\myrhead{\leftmark}


\newpage
\input{tex/WTNotations}



%%%%%%%%%%%%%%
\chapter{Tip-losses: context and challenges }
\input{tex/WTAerodynamicsTipLossesContext}


\chapter{Theories of optimal circulation and tip-losses}
\input{tex/WTAerodynamicsTipLossesTheories}


\chapter{Tip-loss corrections and their implementation}
\input{tex/WTAerodynamicsTipLossesReview}


\chapter{Using a vortex code to investigate tip-losses}
\label{chap:vortexcodetiploss}
\input{tex/VortexCodeTipLoss}


\chapter{Using CFD to investigate tip-losses}
\label{chap:cfdtiploss}
\input{tex/CFDTipLosses}

%%%%%%%%%%%%%%%%
\chapter{Results and comparison of the different codes and approaches}
\label{chap:resultsandcomparisons}
\input{tex/CodeComparisons}

%%%%%%%%%%%%%%%%



\clearpage\newpage
\chapter*{Conclusion}
\addcontentsline{toc}{chapter}{Conclusion}
\myrhead{CONCLUSION}

\paragraph{From theory to the development of a vortex code} A better insight on the origin of tip-losses both historically and physically has been acquired through the first part of this study. Going from simple models with infinite number of blades to more advance ones, the 3D flow characteristics around a wind turbine were progressively put to light. The notion of tip-losses appeared when studying the 3D effects that affect the wind turbine performances. In this study the distinction between flow characteristics close to the rotor and far behind it was emphasized, giving rise to the notions of near-wake and far-wake. The structure of the wake and its dynamics has been presented. Given the important vorticity found in the wake and its concentration to vortex sheets and tip-vortices, a great descriptive and predictive tool was found to be the vortex theory. On one hand, this theory led to the development of important theorem from Munk and Betz which were later complemented by the only two theoretical studies of tip-losses: the work from Prandtl and Goldstein. On the other hand, vortex theory can be applied for numerical methods which leads to the development of different vortex codes. The implementation of a vortex-lattice code was hence chosen to further study tip-losses and account for higher complexity than the theoretical work from the beginning of the 20$^\text{th}$ century.

\paragraph{Historical work on tip-losses} Historically, the concept of tip-losses appeared when looking for the circulation that would lead to the minimum energy loss. By focusing on the theories of Betz and Prandtl, the original notion of tip-losses and its applicability was presented. Nevertheless, it was seen that several interpretations could be given due to the simplicity of Prandtl model compared to a real wake flow. The study of Goldstein's theory which model the wake in a better way also led room for discussion. Results from the theories apply to the far-wake and their relations to the rotor parameters is not straightforward. The pitch of the helix, its rotation and the induced velocities are different than the one at the rotor. Using assumptions of no wake expansion and conservation of circulation relations between far-wake parameters and rotor parameters were established. It has been pointed out that the relations used in contemporary articles are different than the one used originally. As many figures as possible were provided to illustrate the different relations and the focus has been on applying this figure to the wind turbine case to avoid any further confusions. Hopefully with this report, the reader can assimilate the historical work of the German school without going through the original articles which use different notations and conventions. Their reading is nevertheless quite interesting and the author highly recommend the curious reader to discover them. 

\paragraph{Different tip-loss factors} In this study a review of all the existing tip-losses correction, to the author's knowledge, was performed. By using a general formalism for the parameters of the far wake, the theories of Betz, Goldstein and Prandtl were generalized. From these derivations, it was possible to unify all the different tip-loss factor that authors usually refer to as ``Prandtl'' tip-loss factor even-though they use a different formulation than the original author. Hopefully, more clarity on the matter should follow from this analysis. On top of the existence of many tip-loss factor equations, different definitions, interpretations and implementations are found in BEM codes. This study clearly made the distinction between the standard tip-loss factor and the so called ``performance'' tip-loss factor which stand in the line of airfoil coefficients corrections. 


\paragraph{Standard tip-loss factor} The standard tip-loss factor can be defined as a ratio of circulation, like in the original articles of Prandtl and Goldstein, or as a ratio of induced velocities. The later case deserve particular attention. The factor can describe in which proportion the axial induction on the blade is different than the average azimuthal axial induction. In this case though, it should be noted that for a real flow the average azimuthal axial induction goes continuously towards zero in the vicinity of the tip. The axial induction on the blade can take a wide range of values towards the tip due to the formation of the tip-vortex so that the ratio can be hard to define at the very tip of the blade (last 1\%). The advantage of this definition of the tip-loss factor, is that is gives a good definition of the angle of attack and thus a better estimate of the loads on the blade using the blade element theory.  When used in BEM codes a different definition of the tip-loss factor in terms of induced velocities is found. BEM codes are based on momentum theory which models the rotor with an infinite number of blades and hence no tip-losses are present. In that case the tip-loss factor has conceptually the role of reducing the axial induction from momentum theory to ensure is goes towards zero in the vicinity of the tip. In this sense the tip-loss factor would be the ratio of the real azimuthally averaged axial induction to the momentum theory axial induction. In most BEM code implementations though, the tip-loss factor is applied as a momentum (and sometimes mass flow) change. The tip-loss factor then reduces the loads at the tip. In most BEM equations, the distinction between the average axial induction and the induction on the blade is not done. The introduction of the tip-loss factor in the equations has thus the effect of increasing the axial induction locally to the blade. As a result of this, the velocity triangle is correct as in the previous formulation of the tip-loss factor described. The author believe that BEM equations should be rewritten by carefully making the difference between the axial induction on the blade and the azimuthally averaged axial induction. This is done for instance by Shen\cite{shen:2005}. Also, the momentum theory equations should be derived in two steps. The first step consisting in deriving the equations for infinite number of blades. The second step should emphasize the assumption of independence of the different annuli, and hence justify that the axial induction(which is azimuthally averaged by definition), can be changed in each annuli to ensure that it goes towards zero in the vicinity of the tip and hence account for finite number of blades. This should clarify the distinction between $\hat{a}$, $\bar{a}$, and $a_B$ made in this report.

\paragraph{Zero at the tip} From the above discussion and the analysis of this report, the question of the tip-loss factor going to zero at the tip was mentioned. Depending on the meaning of the tip-loss factor different conclusions can be drawn. Clearly, the author believes that the average axial induction goes towards zero in the vicinity of the tip.Nevertheless, this is most likely to happen after the tip accounting for the shear stress at the slip stream boundaries. This is observed from CFD data as well. For this reason, forcing the tip-loss factor to go to zero at the tip might be inappropriate. Also, the flow at the tip is very complex close to the blade. Changes in flow direction might even happen in some situations. Hence if the definition $F=\bar{a}/a_B$ is used then the values at the tip is hard to define and is not likely to be exactly zero. If on the other hand the tip-loss is seen as a circulation factor then one can argue that, even-though the velocity is none-zero from the above discussion, the loads are likely to be reduced to zero due to the pressure balance occurring at the tip.


\paragraph{Performance tip-loss factor} The definition of this performance tip-loss factor and its implementation does not rise as many questions, even though controversy has followed the introduction of Shen's tip-loss factors. This performance tip-loss factor can translate two things. First it can be seen as the results of the pressure equalization at the tip, which should make the loads goes to zero at the very tip. Also, it can model the fact that the airfoil does not perform as in 2D in the tip area due to the presence of the tip-vortex and complex flows occurring. Reasons for this are multiple: radial flow, interaction of the tip-vortex with the boundary layer, separation at the very tip and centrifugal forces, etc. This difference in performance is independent to the other tip-loss factor which affects the relative velocity both in amplitude and in direction. Even if the relative velocity is known perfectly, the 2D airfoil data (in the situations observed) do not apply at the very tip. The performance tip-loss factor is hence defined as the ratio between 3D to 2D airfoil polars. The implementation of such correction is straight-forward in BEM codes. The question that remains is the determination of such factor. Experimental data and CFD data can be used but at this time the imprecision in both methods are quite high at the tip of the blade so no conclusion or persistent model can be derived yet. 


\paragraph{Tip-losses and CFD data} As mentioned in the previous paragraph, the study of the performance tip-loss factor can be done using CFD data. An attempt to derive a model for BEM codes from the available data was done in this study. Nevertheless, imprecisions in CFD and 2D airfoil data leaves the discussion open on the matter. On a different perspective, CFD data was used to analysed tip-losses in terms of axial induction. Interesting results where obtained which matched well the results from the vortex code. This study pointed out the importance of choosing planes as close as possible as the rotor plane in order to assess the tip-loss factor. The definition of the axial induction on the blade was discussed. This definition is difficult due to the spatial extent of the blade and the strong flow fluctuations due to the bound circulation around the blade. Different methods were investigated in this report to assess the axial induction on the blade using CFD data. The method of circular sector with different central angle is preferred by the author for it performs a more physical comparison. On the other hand, only small differences between the methods were found for studying the tip-loss factor only the very tip showed great deviations between the methods.


\paragraph{A new method to investigate tip-losses using a vortex code} In this study, a new method was derived to analyse tip-losses using a vortex code. The idea behind this method is that a unsteady free wake vortex code accounts for finite number of blades, wake expansion and wake roll-up. Due to the simple lifting line formulation, the definition of axial induction on the blade and at the rotor plane is made easy, and so is the derivation of the tip-loss factor. To perform a wide range of characteristic analysis, simulations for different tip-speed ratio, thrust coefficients and different family of circulation curves were run. The derivation of the family of circulation curves was done using B\'ezier curves. The parametrization method derived in this study can be used to characterize typical circulation curve, chord or lift distributions. The author encourages the uses of this method for such applications due to the wide range of curves that can be obtained for only a few parameters. Results from the different vortex code simulations were stored in a database for later use in a BEM code.


\paragraph{Improved BEM code} A new BEM code was implemented using the new tip-loss factors derived by the vortex code. The performances of this new BEM code were compared with the one from the vortex code and from a classical BEM code that uses Glauert tip-loss factor. The new BEM code produces results really similar to the vortex code, and quite different at the tip than the Glauert-BEM code. It is recalled here that the theory of Prandtl is really simplified and moreover does not account for wake expansion and wake roll-up. The vortex code accounts for this, and the fact that the new BEM code reproduces fairly well the results from the vortex code shows that the goal of improving the physical model of the BEM code has been reached. The computational time for the new BEM code is slightly increased, but still small compared to a vortex code or CFD simulation. From its small computational time the new BEM code can be used as a design tools for new blade, and performance prediction of wind turbine.


\paragraph{Further} From the discussions above, it has been seen how difficult the definition of tip-losses was, and how sensitive the implementation of the tip-losses in a BEM code could be. This study intended to give an overview of the existing work on tip-losses and to clarify and distinguish them. Further work is required for the development of BEM equations which clearly distinguish the different axial induction factor introduced in this report. Using either a vortex code, CFD, or experimental data, the link between far-wake and rotor parameters has to be studied in order for the theory of Goldstein to be used more precisely. The use of a vortex code to investigate tip-losses was shown in this study, and the way it could improve BEM code was demonstrated. Methodology similar to the one employed can be derived for further improvements.
 It is also believed that improvement in the development of vortex code can lead to better results in the determination of the tip-loss factor. The investigation of tangential induction factor is also to be considered. Indeed, one can ponder the validity of using a same correction factor for the axial and tangential induction. Discussions from the definition of tip-loss and their implementation will follow this study, making it a challenging but exciting topic.


\newpage
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
\myrhead{ACKNOWLEDGMENTS}
This master's thesis adventure was made possible thanks to the contribution and help of many people, colleagues and friends.
First, I would like to thank Kevin Standish for offering me the opportunity to do my master's thesis at Siemens and to join the dynamic team of Boulder's office. I am grateful for his patience, his perseverance and his generosity and I really appreciated his efforts to make be feel welcomed and part of the team. My thanks are also going to my two supervisors, Kristian Dixon and Mac Gaunaa who regularly followed my work, provided me with valuable advises and guided me with patience. They helped me to focus my study, but discussing with them would arouse my interest and curiosity for the topic and would motivate me to investigate more topics. I hence really appreciated that they offered me this freedom of inquiring and spending time in different directions. During this study I appreciated that several scientists took the time to reply to my emails and questions. Being aware of the few time they have for such considerations, and given the fact that not all my emails received a reply, I am deeply thank full to the one who took this time: Gijs van Kuik, Valery Okulov, Helge Madsen, Christian Bak and Wen Zhong Shen. My experience in Boulder would have not been possible without the help and discussion of Jakob Mann and Jeppe Johansen who oriented me towards Boulder. 


I really appreciated the six months spent amongst Boulder's team and I would like to thank them all for welcoming me in such a nice working environment and for bringing a rich human dimension to this project. They all contributed in a way to this work, and it is difficult not of mention all of them. For this reason, I will mention in the following the ones I thank for their direct contribution to this study: Alonzo Zamora for his collaboration and discussion and for providing all the CFD data this study was based on; Paul Medina for some fun and interesting aerodynamic discussions; Arni Steingrimsson for his interest in converting my code to GPU; Khanh Nguyen for looking over my screen once in while to give me feedback and discuss vortex codes with me; Drew Eisenberg for some programming experience exchanges; and of course Kristian Dixon for supervising me with patience. My stay in Boulder would not have been the same without the spare time activities spent with colleagues and friends. Adding to the above my family together with the not mentioned and forgotten ones, I'd like to thank them all for making this project such a rich adventure. 


\vfill

%\addcontentsline{toc}{chapter}{References}
%\bibliographystyle{plain}
\nocite{larrabee:1983}
\bibliographystyle{acm}
\bibliography{bib/DTUbiblio}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%Annex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newpage
 \appendix
 \myrhead{APPENDICES}
 %  \pagenumbering{Alph}\setcounter{page}{1}
 \addtocontents{toc}{\protect\newpage}
 \addcontentsline{toc}{chapter}{\Large{Appendices}}
 \setcounter{section}{1}

 \par\ 
 \vfill
 \begin{center}\Huge{APPENDICES}
 \vspace{3cm}
 \end{center}
 \vfill
 \par\
 \newpage

 
\chapter{Goldstein's factor - Derivation and computation}
\myrhead{\leftmark}
\input{anx/anxGoldstein}

%%%%%%%%%%%%%%%%%%
\chapter{Implementation and validation of a vortex code}
\label{anx:vortexcode}
\input{tex/VortexCodeImplementation}
\input{tex/VortexCodeValidation}

\input{anx/anxBEM}
\newpage
\input{anx/anxCFD}
\input{anx/anxGermanAbstracts}
\input{anx/anxSourceVortexCode}
% 

\newpage


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%Bibliographie
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newpage
 \addcontentsline{toc}{chapter}{List of figures and tables}
 \listoffigures
 \listoftables


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%Index
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\printindex

\end{document}

